---
title: "Lecture 9 - Tree-Based Methods"
author: "Joseph Haymaker"
date: "11/14/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load all packages
if(!require('pacman')) {
  install.packages('pacman')}
pacman::p_load( ISLR, tidyverse, tree, ramdomForest, ISLR, rpart, rattle, pROC, partykit, gbm)
```

# Tree based methods
+ Text book
  + 5.2: Bootstrap 
  + 8.1: Single trees
  + 8.2: Ensemble methods
      + Bagging
      + Random Forest
      + Boosting

## Part I: regression trees

#### 1) Single Tree
#### 2) Bagging
#### 3) Random Forest (Bagging is a special case)

#### 4) Appendices (all about trees)
#### Appendix I: Testing errors vs. OOB testing erros
#### Appendix II: Effects for different Bootstrap samples on a single tree
#### Appendix III: Comparison among a single tree, bagging and rf

#### 5) Boosting 

## Part II: RandomForest for classifications ( a separate file)

```{r}
library(tree)              # regression/classification trees
library(randomForest)      # to see how a package is evolved to be better: rfNews()
```

#####  Part I: Regression trees
For the purpose of demonstration we use the Baseball data from the book.
+ Goal: predict log salary given the past performance!

```{r}
library(ISLR)
help(Hitters)
data.comp <- na.omit(Hitters)  # For simplicity
dim(data.comp)            # We are keeping 263 players
data.comp$LogSalary <- log(data.comp$Salary)  # add LogSalary
names(data.comp)
data1 <- data.comp[,-19]        # Take Salary out
```


#### 1) A single tree

The idea is to partition the space into boxes. 
  + `Binary split`: Take the best predictor 
  + `Recursive`: Repeat the search in the next half space
  + `The prediction`:  sample mean in each box.
  + (one could use different method: reg for example)

__A single tree__
+ Pros:
  + not a linear model, more flexible
  + take interactions among variables
  + simple interpretation of the prediction
+ Cons:
  + not stable
  + overfitting or not a good prediction

We use it here to illustrate the idea of binary, top-down, recursive trees.

##### Model 0: 

a) Use `CAtBat` alone to see the step function
+ Good start to understand a tree.

```{r}
par(mfrow=c(2,1))

# First optimal split: the node and the estimates
plot(NA, NA, type = "n", xlim = c(0,10000), ylim = c(4, 8), xlab = "CAtBat", ylab = "LogSalary")
points(data1$CAtBat, data1$LogSalary, pch=16)
abline(v=1452, col= "red", lwd=4) 
lines(x = c(0,1452), y = c(5.09, 5.09), lwd=3, col="blue")
lines(x = c(1452, 10000), y = c(6.46, 6.46), lwd=3, col="blue")


fit0.single <- tree(LogSalary~CAtBat, data1)
plot(fit0.single)
text(fit0.single, pretty=0)   # add the split variables

par(mfrow=c(1,1))

fit0.single$frame   # details in each split

summary(fit0.single)
```

```
 var   n        dev     yval splits.cutleft splits.cutright
 1 CAtBat 263 207.153733 5.927222          <1452           >1452
 2 CAtBat 103  36.219531 5.092883           <688            >688
 4 CAtBat  54  18.324402 4.764117         <211.5          >211.5
 8 <leaf>   5  10.550154 5.496828                               
 9 <leaf>  49   4.816009 4.689350                               
 5 <leaf>  49   5.626109 5.455197                               
 3 CAtBat 160  53.076591 6.464327        <1772.5         >1772.5
 6 <leaf>  20   3.593939 6.055648                               
 7 <leaf> 140  45.665082 6.522710      
```

b) Use `CHits` alone to see the step function 

```{r}
par(mfrow=c(2,1))
plot(data1$CHits, data1$LogSalary, pch=16)
fit0.single <- tree(LogSalary~CHits, data1)
plot(fit0.single)
text(fit0.single, pretty=0)  # add the splits
par(mfrow=c(1,1))
fit0.single$frame  # report the nodes
```

```
#       var   n         dev     yval splits.cutleft splits.cutright
# 1   CHits 263 207.1537331 5.927222           <358            >358
# 2   CHits 101  35.2429757 5.083074           <182            >182
# 4   CHits  56  18.3590026 4.771243          <49.5           >49.5
# 8  <leaf>   5   9.3315033 5.660001                               
# 9   CHits  51   4.6908471 4.684110           <132            >132
# 18 <leaf>  34   1.6206148 4.533326                               
# 19 <leaf>  17   0.7511846 4.985678                               
# 5  <leaf>  45   4.6621832 5.471130                               
# 3   CHits 162  55.0688493 6.453511           <450            >450
# 6  <leaf>  16   3.2773291 5.878956                               
# 7  <leaf> 146  45.9308869 6.516476      
```

##### Model 1: Use all the obs'n but only include two variables CHits and CAtBat

```{r}
fit1.single <- tree(LogSalary~CAtBat+CHits, data1) # The order plays no role
fit1.single
names(fit1.single)
fit1.single$frame 
```

```
       var   n         dev     yval splits.cutleft splits.cutright
  1  CAtBat 263 207.1537331 5.927222          <1452           >1452
  2   CHits 103  36.2195305 5.092883           <182            >182
  4   CHits  56  18.3590026 4.771243          <49.5           >49.5
  8  <leaf>   5   9.3315033 5.660001                               
  9   CHits  51   4.6908471 4.684110           <132            >132
  18 <leaf>  34   1.6206148 4.533326                               
  19 <leaf>  17   0.7511846 4.985678                               
  5  <leaf>  47   5.1645460 5.476113                               
  3   CHits 160  53.0765908 6.464327           <669            >669
  6  <leaf>  50   9.5004560 6.213632                               
  7  <leaf> 110  39.0053795 6.578279   
```

```{r}
plot(fit1.single)
text(fit1.single, pretty=0) 
```
     
+ `pretty`=0 only affect the categorical var's. The names will be shown.
+ It has 6 terminal nodes. That means we partition `CAtBat` and `Chits` into six boxes. 
+ The predicted values are the sample means in each box.

```{r}
fit1.single.result <- summary(fit1.single)
names(fit1.single.result)

fit1.single.result$dev # dev=RSS. 
fit1.single.result$size  # number of nodes
fit1.single.result$df  # n- number of nodes
```


+ RSS=65.37. It should be same as the sum of RSS at each `terminal node``
+ 9.332+1.621+0.751+5.165+9.500+39.010=65.38 (should be the same)

```{r}
yhat <- predict(fit1.single, data1)  # only size=6 many predicted values
RSS.tree <- sum((data1$LogSalary-yhat)^2 )   # another way to get RSS
RSS.tree  #or 
sum((fit1.single.result$res)^2)
# [1] 65.37368

plot(data1$LogSalary, yhat, pch=16, col="blue",
     xlab="LogSal",
     ylab="Yhat")
```

As we already knew there are only __6 predicted values__ being used.

__QUESTION__: How does the above tree perform comparing with our old friend lm in terms of the in sample errors?

```{r}
fit.lm <- summary(lm(LogSalary~CAtBat+CHits, data1))
RSS.lm <- (263-2)*(fit.lm$sigma)^2
RSS.lm ## Oppps much worse than even a single tree. 
# [1] 127.2509   RSS.tree=65, in sample RSS is much smaller from the above tree.
```

##### An alternative tree (I like its output)

```{r}
library(rpart)
fit.single.rp <- rpart(LogSalary~CAtBat+CHits, data1, minsplit=20, cp=.009)
fit.single.rp 
plot(fit.single.rp)
text(fit.single.rp, pretty = TRUE)   # plot method 1
summary(fit.single.rp)
```

`fancyRpartPlot` does seem to work with the current R version


```{r}
#install.packages('rattle')
#install.packages("RGtk2")
# library(RGtk2)
# library(rattle)  # rattle is needed to produce a neat tree coupled with rpart
library(partykit) # as.party

#fancyRpartPlot(fit.single.rp) # plot method 2  (can't get it work)
plot(as.party(fit.single.rp), main="Final Tree with Rpart") # method 3
```

The plots are only useful for a small tree!

##### Model 2:  use all the predictors

```{r}
fit1.single.full <- tree(LogSalary~., data1) 
```

We can control how large a tree we want to build: 

  + `control=tree.control( minsize = 6, mindev = 0.05))`      
  + the default `tree.control(nobs, mincut = 5, minsize = 10, mindev = 0.01)`
  + `mincut`: min number of obs’n to be included in a child
  + `minsize`: number of end nodes
  + `mindev`: the dev needs to be reduced by mindev fold within the branch: `dev_new < mindev * dev_parent`

Here is the tree:

```{r}
plot(fit1.single.full)  
text(fit1.single.full, pretty=0)
fit1.single.full$frame

fit1.single.full.s <- summary(fit1.single.full)
names(fit1.single.full.s)
names(fit1.single.full)

fit1.single.full.s$dev  # RSS=43.03
sum((fit1.single.full.s$residuals)^2) # to check the dev is the RSS

fit1.single.full.s$used # Var's included 
```

7 variables are used:

+ `CAtBat`, `CHits`, `AtBat`, `CRuns`, `Hits`, `Walks`, `CRBI`
+ There are 9 terminal nodes

```{r}
fit1.single.full.s$size

RSS.lm <- (263-8)*((summary(lm(LogSalary~CAtBat+CHits+AtBat+CRuns+Hits+Walks+CRBI, data1)))$sigma)^2
RSS.lm   # Still pretty big 101.5803
```

__Question__: do you expect that the testing `RSS(lm)` > `RSS(single tree)` ?

### 2)  Bagging: general method

```
Algorithm:
  1) Take B many bootstrap samples
  2) One tree for each B sample
  3) The final predictor = Average of all B  trees
  4) It is a special case for Random Forest
```
+ __Pros__: Reduce the var while maintain similar bias. 
+ __Cons__: The trees are correlated (on higher level). 

### 3) Random Forest

```
Algorithm:
  1) Take B many bootstrap samples
  2) Build a deep random tree for each Bootstrap sample by
    * Split only m (mtry) randomly chosen predictors at each split
  3) Bag all the random trees by taking average => prediction of y given x_1,... x_p
  4) Use Out of Bag testing errors to tune mtry!!!!! 
```

+ __Pros__: Decorrelate the trees - reduce var more
+ __Cons__: Tuning parameter `m` is introduced 
  + `m` too small - miss important var's
  + `m` too large - more cor between trees

Remark: 
1) nodesize:5 for reg's and 1 for classifications
2) when mtry=p (=19), randomForest gives us bagging estimates.

```{r}
library(randomForest)
```

i) For a fixed mtry

```{r}
fit.rf <- randomForest(LogSalary~., data1, mtry=6, ntree=100)
plot(fit.rf)  # repeat this and you see variability in the plots why so???
str(fit.rf)
```

default settings: 
+ `mtry`=p/3, (sqrt(p) in classification tree)
+ Bootstrap size `B`=`ntree`=500 
+ when `mtry`=`p`=19, randomForest gives us bagging estimates.
+ nodesize:5 for reg's, and 1 for classifications. 

Based on the 100 random trees, we get hat y= ave (all 100 trees).
How does it perform? 

```{r}
yhat <- predict(fit.rf, data1)   # predicted values we use 

plot(data1$LogSalary, yhat, pch=16,  # add a 45 degree line:looks very good!
     main="Y vs. Predicted Y", col="blue")
abline(0, 1, lwd=5, col="red")

mse.train <- mean((data1$LogSalary-yhat)^2)  # RSS= sum((data1$LogSalary-yhat)^2)
mse.train      # Training mse approx. .03!!!
mse.oob <- mean((data1$LogSalary-fit.rf$predicted)^2)  #OOB testing error 
mse.oob        # about .18  
```


ii) Zoom in to a RF estimate

```{r}
fit.rf <- randomForest(LogSalary~., data1, mtry=10, ntree=100)
#fit.rf <- randomForest(LogSalary~., data1, mtry=10, ntree=2)
str(fit.rf)
```


a) Where is cross validation testing error? It is cleverly replaced by OOB mse

```{r}
par(mfrow=c(2,1))
plot(fit.rf$mse, xlab="number of trees", col="blue",
     ylab="ave mse up to i many trees using OOB predicted",
     pch=16) # We only need about 100 trees for this 
# We get the above plot by 
plot(fit.rf, type="p", pch=16,col="blue" )
par(mfrow=c(1,1))
```

b) OOB's are used

```{r}
fit.rf$oob.times  # how many times each obs'n belong to OOB. We expect to see 1/e=1/3 (.37)

#   fit.rf$mse is OOB testing errors = mse of only using the OOB fitted values up to tree number
#   fit.rf$predicted is obtained only using the OOB obs'n 

# fit.rf$predicted[]   # predicted values based on all OOB values. we use this to estimate the mse which is approx. the testing error
# predict(fit.rf, data1)  # predicted values based on the RF. this will be sued for predictions.

plot(fit.rf$predicted,predict(fit.rf, data1), pch=16, col="blue",
     main="RF prediction vs. OOB prediction",
     xlab="Fitted using OOB only",
     ylab="Fitted using RF")
abline(0, 1, col="red", lwd=5)   # They differ but not by too much.
```


b) Ready to tune `mtry` and `B`=number of the trees in the bag

i) `ntree` effect: given mtry and ntree, we see the effect of ntree first

```{r}
fit.rf <- randomForest(LogSalary~., data1, mtry=10, ntree=500)
plot(fit.rf, col="red", pch=16, type="p", main="default plot, OOB errors")
```

We may need 250 trees to settle the OOB testing errors

ii) The effect of `mtry`: the number of random split at each leaf

Now we fix `ntree`=250, We only want to compare the OOB mse[250] to see the mtry effects.

Here we loop `mtry` from 1 to 19 and return the testing OOB errors

```{r}
par(mfrow=c(3,1))
rf.error.p <- 1:19  # set up a vector of length 19
for (p in 1:19)  # repeat the following code inside { } 19 times
{
  fit.rf <- randomForest(LogSalary~., data1, mtry=p, ntree=250)  
  #plot(fit.rf, col= p, lwd = 3)
  rf.error.p[p] <- fit.rf$mse[250]  # collecting oob mse based on 250 trees
}
rf.error.p   # oob mse returned: should be a vector of 19

plot(1:19, rf.error.p, pch=16,
     xlab="mtry",
     ylab="mse of mtry") 
```

Run above loop a few time, it is not very stable. 

__Notice__
1) mtry = 1 is clearly not a good choice.
2) The recommended mtry for reg trees are mtry=p/3=19/3 about 6 or 7. Seems to agree with this example. Are you convinced with p/3?

We should treat `mtry` to be a tuning parameter!!!!!!!!  

c) The final fit: we take `mtry`=6

```{r}
fit.rf.final <- randomForest(LogSalary~., data1, mtry=6, ntree=250)
plot(fit.rf.final)
fit.rf.final$mse[250] # approx is same as the testing error
```

d) Prediction

```{r}
person <- data1[1, ]   # Let's predict rownames(data1)[1]: "-Alan Ashby"
fit.person <- predict(fit.rf.final, person)
fit.person # the fitted salary in log scale is 6.196343 (may not be the same each time we run the rf, why?)
# -Alan Ashby 
# 6.170148 

# Alan Ashby's true log sal 
data1$LogSalary[1]
# [1] 6.163315
```


End of randomForest in regressions.  

### 4) Appendices

### Appendix I: Testing errors vs. OOB testing erros

We could also get testing errors w/o using OOB idea
The following code verify that OOB errors seem to be a good estimate of testing errors.

```{r}
n <- nrow(data1)
#set.seed(1)
train.index <- sample(n, n*3/4) # we use about 3/4 of the subjects as the training data.
train.index
data.train <- data1[train.index,]
data.test <- data1[-train.index, ]

fit.rf.train <- randomForest(LogSalary~., data.train,  mtry=6, ntree=500) 
```

This will output the oob errors

To get testing error:

```{r}
fit.rf.testing <- randomForest(LogSalary~., data.train, xtest=data.test[, -20], 
                          ytest=data.test[,20], mtry=6, ntree=500)
# fit.rf.testing$mse
```


This will output the testing errors directly

```{r}
plot(fit.rf.testing$mse)

plot(fit.rf.train) # when xtest and ytest are given, the output will be for testing data, like y, mse, etc
```

The testing errors seem to agree with what we found using OOB errors.

Let's put testing errors and OOB errors together

```{r}
plot(1:500, fit.rf.testing$mse, col="red", pch=16,
     xlab="number of trees",
     ylab="mse",
     main="mse's of RF: blue=oob errors, red=testing errors")
points(1:500, fit.rf.train$mse, col="blue", pch=16)
```

OOB errors seem to do ok to estimate the testing errors!!!
Not bad at all.....


### Appendix II: Effects for different Bootstrap samples on a single tree

+ Trees built by resampling the data through BOOTSTRAP method
+ Take a bootstrap sample and exam the tree built by the sample
  i) About 2/3 of the original subjects are chosen in each Bootstrap sample
  ii) The trees are different
  ii) mse changes a lot

```{r}
RSS <- 0  # initial values
n.unique <- 0
n <- nrow(data1); K <- 20
# for (i in 1:K)
# {
#   index1 <- sample(n, n, replace=TRUE)   
#   Sample1 <- data1[index1, ]               # Take a bootstrap sample
#   fit1.boot <- tree(LogSalary~., Sample1)  # Get a tree fit
#   RSS[i] <- summary(fit1.boot)$dev  # output RSS for each bootstrap tree
#   plot(fit1.boot, 
#        main="Trees with a Bootstrap sample") 
#   text(fit1.boot, pretty=0)
#   n.unique[i] <- length(unique(index1))
#   Sys.sleep(2)                        # Pause for 2 seconds before running for next round
# }
# 
# hist(RSS, breaks=30, 
#      col="blue",
#      main="RSS from different Bootstrap trees")
# 
# 
# hist(n.unique, breaks=30,
#      col="red", 
#      main="number of unique subjects included in each Bootstrap sample")
# 
# hist(n-n.unique, breaks=30,
#      col="green", 
#      main="number of OOB subjects not included in each Bootstrap sample")
```

```{r}
RSS <- 0  # initial values
n.unique <- 0
n <- nrow(data1); K <- 20
# for (i in 1:K)
# {
#   index1 <- sample(n, n, replace=TRUE)   
#   Sample1 <- data1[index1, ]               # Take a bootstrap sample
#   fit1.boot <- tree(LogSalary~., Sample1)  # Get a tree fit
#   RSS[i] <- summary(fit1.boot)$dev  # output RSS for each bootstrap tree
#   plot(fit1.boot, 
#        main="Trees with a Bootstrap sample") 
#   text(fit1.boot, pretty=0)
#   n.unique[i] <- length(unique(index1))
#   Sys.sleep(2)                        # Pause for 2 seconds before running for next round
# }
# 
# hist(RSS, breaks=30, 
#      col="blue",
#      main="RSS from different Bootstrap trees")
# 
# 
# hist(n.unique, breaks=30,
#      col="red", 
#      main="number of unique subjects included in each Bootstrap sample")
# 
# hist(n-n.unique, breaks=30,
#      col="green", 
#      main="number of OOB subjects not included in each Bootstrap sample")
```

### Appendix III: Compare a single tree, a bag with 100 trees and a RF with mtry=10, 100 trees

A single tree

```{r}
fit.single <- randomForest(LogSalary~., data1, mtry=19, ntree=1)
```

Bagging with 100 Bootstrap samples

```{r}
fit.bagging <- randomForest(LogSalary~., data1, mtry=19, ntree=100)
```

Random Forest

```{r}
fit.rf <- randomForest(LogSalary~., data1, mtry=6, ntree=100)
```


Compare performance among the three methods graphically:

```{r}
par(mfrow=c(3,1))
ylim.0 <- c(3.5, 7.5)
plot(data1$LogSalary, predict(fit.single, data1), pch=16,  # add a 45 degree line:looks very good!
     ylim=ylim.0,
     main="A single tree", col="blue")
abline(0, 1, lwd=5, col="red")

plot(data1$LogSalary, predict(fit.bagging), pch=16,  # add a 45 degree line:looks very good!
     ylim=ylim.0,
     main="A bagging tree, 100 trees", col="blue")
abline(0, 1, lwd=5, col="red")

plot(data1$LogSalary, predict(fit.rf), pch=16,  # add a 45 degree line:looks very good!
     ylim=ylim.0,
     main="A RF tree, mtry=10, 100 trees", col="blue")
abline(0, 1, lwd=5, col="red")
par(mfrow=c(1,1))
```

Testing errors through OOB among the three:

```{r}
c(fit.single$mse, fit.bagging$mse[100], fit.rf$mse[100])
```

RF is the best (??) but it should have performed much better than a bagging estimate.
Are the three mse's comparable????

#### 5) Boosting

```{r}
library(gbm) # boosting machine
data1 <- data1[sample(nrow(data1)), ] # make sure the row number plays no row!
```


i) Boosting in action

```{r}
ntree <- 200
fit.boost <- gbm(LogSalary ~., data = data1, distribution = "gaussian", n.trees = ntree, interaction.depth = 2,
                 train.fraction = .7)
```

__arguments:__

+ `distribution` = "gaussian" for continuous response and "bernoulli"/"multinomial" for two/more two/more classes.
+ `interation.depth` = depth of the tree, the number of splits.  d=1 (default) is a stump tree, i.e. only one split
+ `n.trees` = number of the trees to be updated, default=100
+ `train.fraction` = training set, the error will be testing error with the remaining data.
 __Caution__: the first train.fraction of the data will be used as a training set
    + Make sure the data is in a random order.                    
+ `bag.fraction` = fraction of the training data, fraction of the data used for training the next tree
               default set at .5
+ `shrinkage` = 0.001 set as default.

##### Output

```{r}
names(fit.boost)
fit.boost$fit # hat y
fit.boost$train.error # training errors 
fit.boost$valid.error # testing errors if train.fraction is given

yhat <- predict(fit.boost, newdata = data1, n.trees = ntree) # prediction, in this case it is the same as fit.boost$fit
```

ii) Tune number of trees needed by minimizing the testing errors

Investigate the testing/training errors vs. number of iterations
We may use:
`gbm.perf(object, plot.it = TRUE, method="test")`

__Remember that the training data will be the first 80% rows.__

```{r}
ntree <- 20000
fit.boost <- gbm(LogSalary~., data = data1, distribution = "gaussian", n.trees = ntree, interaction.depth = 2,
                 train.fraction = .8)
gbm.perf(fit.boost, method ="test") 
```

Output the optimal number of trees by min the testing errors.
+ red: `fit.boost$valid.error` (testing errors)
+ black: `fit.boost$train.error` (training errors)

Notice the testing errors are smaller than that of training error for many iterations.

The above plots are same as: 

```{r}
plot(fit.boost$valid.error, col= "red", 
     main="red: testing errors, black: training errors", 
     ylim = c(0, .9))
points(fit.boost$train.error, col="black")
```

We now use the optimal number of trees as our final prediction function

To match with the gbm() training and testing data, we set:

```{r}
n.t <- floor(.8*263)
data.train <- data1[1:n.t, ]   # n.t <- floor(.8*263)
data.test <- data1[-(1:n.t), ]

B <- gbm.perf(fit.boost, method ="test") # optimal number of trees
yhat <- predict(fit.boost, newdata = data.test, n.trees = gbm.perf(fit.boost, method ="test") )

fit.boost.test <- mean(yhat-data.test$LogSalary)^2  # the testing error for boosting
```

iii)  Lastly let's compare the testing errors between random forest and boosting

```{r}
fit.rf.testing <- randomForest(LogSalary~., data.train, xtest=data.test[, -20], 
                                         ytest=data.test[,20], mtry=6, ntree=500)
fit.rf.testing$mse[500]   # testing error
```

output the testing errors for boosting and rf

```{r}
print(c(fit.boost.test,fit.rf.testing$mse[500] )) ### wowwww, who's a winner????
```











