---
title: "Midterm"
graphics: yes
date: "Nov 1, 2016, 6 - 8 PM"
output:
  word_document:
    toc: yes
    toc_depth: 2
  pdf_document:
    keep_tex: yes
    toc: yes
  html_document:
    number_sections: yes
    self_contained: no
    toc: yes
    toc_depth: 2
subtitle: STAT 471/571/701
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=TRUE, fig.width=6,  fig.height=5, fig.align='left', dev = 'pdf')
```




**Name:**  ______________________________ 

# Instructions

This exam requires you to use R. It is completely open book/notes. Write your answers in a Word (`.docx`) or R Markdown(`.rmd`) format. It is not a good idea to use `.rmd` if you are not familiar with it yet. Always attach the plots or R output where needed. On the other hand, hide/control unnecessary code or output for knitr users. If you have trouble to format the plots, don’t worry about it. We are not looking for especially pretty solutions, but rather to see if you could make sense out of the data using R.

**Data needed**

Available on Canvas: `/canvas/Data/Newsweek_graduation_subset.csv`

**Electronic Submission**

In the ‘Assignments’ section of Canvas, go to the ‘Midterm’ assignment and you will be able to upload your completed files. If you used knitr, upload your `.rmd` and a compiled file (e.g. `.pdf`) for knitr users. If you didn't, attach your writeup with R code attached as an appendix for non-knitr users. Make it clear which parts of your code correspond with which parts of your analysis. **A write up inside R with comments is not acceptable!**

***The submisison assignment will close at 8:10 pm.***
 
**FAQ**

* If you have trouble uploading your files, email them to `lzhao@wharton.upenn.edu`.
* As always skip any part you have trouble with and you may come back to finish it if you have time. Ask one of us for help if you are stuck somewhere.
* I have written a few lines introducing the data, helping out with data downloading, and extracting a subset.
* None of the questions are related, so skip any part you have trouble with and finish it later.

Graduation rate is a key measurement of success among colleges.  Identifying what affects graduation rate will help schools to improve this important outcome for their students.

Newsweek Magazine collected a data set about college outcomes in 1995. The data includes graduation rates for US colleges, various metrics describing the student body and faculty members, as well as several other descriptors of college life. After intensive sanitizing work, we cleaned the original data set and created some additional useful variables. We will work with this revised version of the data, available in `/canvas/Data/Newsweek_graduation_subset.csv`.

Out of all variables, we are in particularly interested in the following 13 variables. Note that most of them are self-explanatory. 

`Grad.rate:` Graduation Rate

`Name:` Name of the school

`State:` Location

`Schooltype:` 1 = Public, 2 = Private

`All.test.std:` A standardized summary of SAT, ACT scores for admitted students

`App.accept:` # of accepted applicants that year

`Acc.Rate:` The percent of applicants for admission, who were accepted by the college

`Pct.Yield:` The percent of accepted students, who actually enrolled

`Total.students:` Total number of students in the school

`Student.Faculty:` Student-faculty ratio

`Pct.fac.degree:` Percentage of faculty with highest degrees

`In.Tuition:` In state tuition

`Room.board:` Room and board costs

We first do some basic data exploration to familiarize ourselves with the data set. We then use this data to develop statistical models for exploring factors related to the graduation rate and for predicting graduation rate.

\newpage

# Question 1: The College Dropout

Read the Newsweek dataset into R; here are a few lines to aid you. Notice that Penn is in **row 820**. 

```{r, message=FALSE, warning=FALSE}
# Required libraries - you may add more packages as desired
library(leaps)  # regsubsets() for model selection
library(car)    # Anova()
library(glmnet) # glmnet() and cv.glmnet()
```

```{r, data prep, results='hide'}
rm(list=ls()) # Remove all the existing variables

college_data <- read.csv("Newsweek_graduation_subset.csv")
str(college_data)
college_data$Schooltype <- as.factor(college_data$Schooltype)
```

We also identify and isolate Penn within this dataset, to use for prediction purposes later.

```{r}
penn_loc <- which(college_data$Name == "University of Pennsylvania")  # identify Penn
Penn <- college_data[penn_loc, ]
```

**a)** How many colleges are included in this dataset? How many variables are there in this data set? List all the variable names. Apart from continuous variables indicate which variables are categorical variables. Make sure they are treated as factors in `college_data`.

**b)** Which school has the highest graduation rate, and what is that rate? Which school has the lowest graduation rate, and what is that rate? What was Penn's graduation rate in 1995? What is the mean graduation rate, across all schools? *Do not actually include the summary statistics.*

**c)** Write a very short (max 3 sentences) summary about the distribution of graduation rate, and provide a histogram of graduation rate. 

**Assume all linear model assumptions are met in the following analyses.**

# Question 2: School Type and State vs Graduation Rate 


**a)** Make a back to back boxplot of graduation rate vs school type. Does one type seems to have higher a graduation rate compared to the other type? Write a short (max 3 sentences) summary of this finding. Does that agree with your intuition about private schools (`Schooltype = 2`) vs. public schools (`Schooltype = 1`)?
	
**b)** `fit1`: `Grad.rate` vs. `Schooltype`
	
Perform a test to determine if the mean `Grad.rate` between the two school types is different at .01 level. Which type has a higher `Grad.rate`? Produce a 95% confidence interval for the mean difference. 

**c)** `fit1.1`: `Grad.rate` vs. `State` 
 
Can we prove that the mean graduation rates are different, at the 0.01 level, among all the states? Which state appears to have the highest graduation rate, and which state appears to have the lowest graduation rate? Note that `AK`/Alaska is the base case in this analysis.

**d)** `fit1.2`: `Grad.rate` vs. `Schooltype` and `State`
	
Controlling for school type, is the school's state a useful factor at the .01 level?

**e)**  Write a few (max 3 lines) sentences to summarize your findings in Question 2. 


# Question 3: Faculty effects

Consider the variable `Pct.fac.degree`, which summarizes the percent of faculty members who hold higher education degrees. 

Construct `fit2`: `Grad.rate` vs. `Pct.fac.degree`

**a)** Report the summary of your linear model. Is `Pct.fac.degree` a significant variable in this model at .05 level? How does `Pct.fac.degree` affect `Grad.rate`?

**b)** Make a scatter plot with $y$ = `Grad.rate` and $x$ = `Pct.fac.degree`. Overlay `fit2` onto the plot.

Construct `fit2.1`: `Grad.rate` vs. `Pct.fac.degree + All.test.std`

**c)** Is `Pct.fac.degree` still a significant variable in this model at the .05 level?

**d)** Interpret the coefficient of `Pct.fac.degree` in `fit2.1`.
  
**e)** Why might the two beta's for `Pct.fac.degree` differ? 


# Question 4: Parsimonious models (all subsets)    (We could take this question out?????)

Construct `fit3`: a model with all available sensible variables 

Based on fit3, answer the following questions:
 
**a)** Is `State` a significant variable at .01 level after controlling for all other variables in the model? Provide an appropriate test. 
 
**b)** If you were to kick one variable out from this model such that the resulting model would have the smallest possible RSS, which variable would you choose and why? (State will be considered as one variable.)
 
Construct `fit4`: a parsimonious model, using exhaustive subset search. Remove `State` from the data under consideration but include all other variables.
 
**c)** Show the Cp plot and also show the BIC plot. Based on the two plots which model size is most desirable to choose. Why?
 
**d)** Regardless of your answer in c), report the size 4 variable chosen by regsubsets, list those top four variables. To save time we will not pursue further.
 
# Question 5: Parsimonious models (LASSO) 
 
Use LASSO for model selection, again making sure to do so without including the `State` variable in the LASSO process.

**a)** Run `cv.glmnet()` with `set.seed(12)`. Plot `cmv` vs. `lambda`. 

**b)** What is the `lambda.1se` value? Under the `lambda.1se` criterion, list the non-zero variables returned. 

**c)** `fit5`: Run OLS with all the variables returned from part b), *and include State.* Are all the variables included here significant at the .01 level? If not, perform backward elimination (manually) until all the p-values for the remaining variables are less than .01. Show your model building process and report the final LS equations. *Note:* for this problem, force State as a whole, into the final model, i.e., do not remove State.

# Question 6: Graduation Evaluation

Independent from Question 5 - assume that we've decided to use `fit6` as our final model.

`fit6`: `Grad.rate` ~ `State + Schooltype + All.test.std`

**a)** Are all three variables significant at the .01 level? 

**b)** Provide the residual plot.

**c)** Provide the qqnorm plot of the residuals.

**d)** Are the linear model assumptions met? 
  
**e)** Amy Gutmann's asked you to present a (short - max 6 sentences) executive summary of your findings to her, and suggest ways to improve Penn's graduation rate. Focus in particular on how each factor in `fit6` affects `Grad.Rate`.
 
**f)** Finally, using `fit6`, provide a 95% prediction interval for Penn's graduation rate. Based on Penn's actual graduation rate, how would you consider our prediction performance?
 
 
# Question 7: Freedom of the Press

Newsweek did a great job of collecting granular data, but some schools are unhappy with their exact graduation figures being reported. They've lobbied Newsweek's publisher to report only whether a school's graduation rate is either High (`Grad.rate >= 70`) or Low (`Grad.rate < 70`), and the "journalists" have acquiesced to their corporate overlords. From now on, the only graduation rate  available to you is in that high/low form.

**a)** Create a new categorical variable `Grad.rate.2` in `college_data` that fits the new Newsweek data specification. What proportion of the schools are categorized as "High Graduation", that is, `Grad.rate.2 == "1"`?

**b)**  How well can we predict `Grad.rate.2`, with only three variables: `State`, `Schooltype` and `All.test.std`. Run a logistic regression of `Grad.rate.2` vs. `State`, `Schooltype` and  `All.test.std`. Is every variable significant at .01 level, while controlling for the other 2 variables?

**c)** Let us fix our classification threshold to 0.5, that is, we will classify each school to be "High Graduation" if the estimated probability of being "High Graduation" is greater than 0.5. Under this framework, what is the in-sample mis-classification error? Show your work.




